Protein-intrinsic properties and context-dependent effects regulate pioneer-factor binding and function

### 2023-10-20 manuscript/manuscript_sections/_06_Figure_Legends.qmd
update analysis and MS based on reviewer comments

**Figure 1: Ectopically expressed pioneer factors open chromatin and activate transcription. a,b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for wild-type (WT) cells lacking Zld or Grh expression.
**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h)**.
**i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions (shaded area) where changes in chromatin accessibility (shaded region) are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
**j,l,** Proportion of class I, II or III regions that is proximal to a gene that is differentially expressed (DE) upon Zld **(j)** or Grh **(l)** expression.
Promoter-proximal and promoter-distal binding sites are shown separately.
The grey dotted line indicates the percentage of all binding sites that are proximal to a differentially expressed gene.

**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites. a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.

**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
Zld, Grh, and Twi ChIP-seq data were generated in this study.
Remaining ChIP-seq datasets were previously published (see Supplementary table 2)[@Henriques2018; @Philip2015; @Tettey2019].
**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
Logos plot of the canonical motif are shown alongside the heatmap.
Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency within regulatory elements.
**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
Motif scores were determined by comparing each identified motif instance to the canonical motif.
For all boxplots, line shows the median, boxes extend from the 25th to the 75th percentile, and whiskers show 1.5 × the interquartile range.
Outlier points beyond the range of the whiskers are shown individually.

**Figure 4: Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in wild-type S2 cells are also shown.
**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
Zld, Grh, and Twi ChIP-seq data are z-score normalized.
CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).

**Figure 5: Pioneer-factor binding and opening of closed chromatin is concentration dependent. a,e,i,** Boxplots of ChIP-seq signal at class I-III regions for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
**b,f,j,** Boxplots of ATAC-seq signal at class I-III regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
**c,g,k,** Distribution of class II and III peaks when defined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
Y-axis indicates the total number of peaks called at each concentration.
Colors indicate the proportion of those peaks that are defined as class II, or III.
**d,h,l,** Venn diagrams showing overlap of peaks called at physiological or supraphysiological expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.

**Figure 6: Regions outside the DNA-binding domain are required for pioneer-factor function. a,b,** Genome browser tracks showing example of class III regions where expression of the DBD alone is insufficient for robust binding and chromatin opening.
Adjacent class I regions are shown where the DBD alone can bind.
Tracks shown for Zld **(a)** and Grh **(b)**.
Class III regions are highlighted with gray shading.
**c,e,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(e)**.
**d,f,** Volcano plots for ATAC-seq data showing the absence of changes to chromatin accessibility upon expression of Zld **(d)** or Grh DBD **(f)**.
**g,** Model for the regulation of PFs.
A high level of chromatin occupancy is required for PFs to initiate chromatin opening.
Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors (outlined in blue).

**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations. a,** Schematic of generation of stable cell lines and induction of protein expression.
**b,** mRNA levels of *zld* and *grh* in S2 cells.
Top histogram shows the distribution of mRNA levels for all *Drosophila* genes.
Vertical dashed line indicates a log~2~ RPKM value of 0 as a threshold for considering a gene to be expressed.
**c-d,** Immunoblots showing titration of Zld **(c)** or Grh **(d)** protein levels in stable cell lines.
Two independently generated cell lines are shown and compared to 2-3 hours (H) embryos.
60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hours embryos.
Black arrowheads indicate Zld **(c)** or Grh **(d)**.
Gray arrowheads indicate background bands used to assess loading.
**e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells.

@@

**Extended Data Figure 8: Loss of H3K27me3 in tazemetostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
Tubulin levels are shown as a loading control.
**b,** Heatmap showing specificity of anti-H327me3 antibody in CUT&RUN reactions.
A panel of barcoded spike-in nucleosomes bearing different modifications was added to each CUT&RUN reaction (see methods).
For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.

**Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites. a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld **(d)**, Grh **(e)**, or Twi **(f)** when expressed at varying concentrations.
**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.

@@

**Supplementary table 2: Previously published ChIP-seq datasets that were analyzed in this work.** For previously published ChIP-seq datasets, this table provides the target of immunoprecipitation, the cell type and/or developmental stage, the Gene Expression Omnibus accession number (GSE), and the PubMed ID for the study in which the dataset was originally published..

>>>>

**Figure 1: Ectopically expressed pioneer factors open chromatin and activate transcription. a,b,** Genome browser tracks showing examples of individual class I, II or III regions for Zld **(a)** or Grh **(b)**.
ChIP seq and ATAC-seq signal are shown for Zld- or Grh-expressing cells along with ATAC-seq signal for wild-type (WT) cells lacking Zld or Grh expression.
**c,f,** Pie charts showing the distribution of class I, II, and III binding sites for Zld **(c)** or Grh **(f)**.
**d,g,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions for Zld **(d)** or Grh **(g)**.
**e,h,** Heatmaps showing ChIP-seq and ATAC-seq signal at class II and III regions for Zld **(e)** or Grh **(h)**.
**i,k,** Example genome browser tracks showing ChIP-seq, ATAC-seq and RNA-seq signal at class III regions (shaded area) where changes in chromatin accessibility (shaded region) are associated with changes in gene expression for Zld **(i)** or Grh **(k)**.
**j,l,** Proportion of class I, II or III regions that is proximal to a gene that is differentially expressed (DE) upon Zld **(j)** or Grh **(l)** expression.
Promoter-proximal and promoter-distal binding sites are shown separately.
The grey dotted line indicates the percentage of all binding sites that are proximal to a differentially expressed gene.
Expression was induced for 48 hours.

**Figure 2: Twist binds closed chromatin extensively and drives accessibility at a limited number of sites. a,** Genome browser tracks showing examples of individual class I, II or III regions **b,** Pie charts showing the distribution of class I, II, and III binding sites.
**c,** The proportion of class I, II or III binding sites that occur at promoters (-500 to +100 bp around transcription start site) or promoter distal regions.
**d,** Heatmap showing ChIP-seq and ATAC-seq signal at class II and III regions.
Expression was induced for 48 hours.

**Figure 3: Motif content shapes pioneer-factor activity. a,** Venn diagram showing overlap between class I regions for Zld, Grh and Twi.
**b,** Heatmap of ChIP-seq signal at class I sites that overlap between Zld, Grh, and Twi.
Cobound indicates regions bound by all three factors.
Not cobound indicates regions bound by any combination of two factors.
Zld, Grh, and Twi ChIP-seq data were generated in this study.
Remaining ChIP-seq datasets were previously published (see Supplementary table 2)[@Henriques2018; @Philip2015; @Tettey2019].
**c-e,** Heatmaps showing the percentage of regions containing a canonical Zld **(c)**, Grh **(d)**, or Twi motif **(e)**.
Logos plot of the canonical motif are shown alongside the heatmap.
Percentages are shown for class I, II or III binding sites, or all wild-type ATAC-seq peaks as a control for the background motif frequency within regulatory elements.
**f,h,j,** Boxplots of the average number of motifs per peak in class I, II, or III regions for Zld **(f)**, Grh **(h)**, or Twi **(j)**.
**g,i,k,** Boxplots of the average motif log-odds score for class I, II or III regions for Zld **(g)**, Grh **(i)**, or Twi **(k)**.
Motif scores were determined by comparing each identified motif instance to the canonical motif.
For all boxplots, line shows the median, boxes extend from the 25th to the 75th percentile, and whiskers show 1.5 × the interquartile range.
Outlier points beyond the range of the whiskers are shown individually.

**Figure 4: Many endogenous binding sites are resistant to ectopic pioneer-factor binding. a-c,** Heatmaps comparing ChIP-seq signal in across different tissues for Zld **(a)**, Grh **(b)**, or Twi **(c)**.
ChIP-seq data for repressive histone modifications H3K27me3 [@Jain2020] and H3K9me3[@Eastwood2021] in wild-type S2 cells are also shown.
**d-f,** Heatmaps comparing DMSO or tazemetostat (taz)-treated cells for Zld **(d)**, Grh **(e)**, or Twi **(f)**.
Zld, Grh, and Twi ChIP-seq data are z-score normalized.
CUT&RUN data for H3K27me3 is spike-in normalized using barcoded nucleosomes (see methods).
Expression was induced for 48 hours.

**Figure 5: Pioneer-factor binding and opening of closed chromatin is concentration dependent. a,e,i,** Boxplots of ChIP-seq signal at class I-III regions previously identified for Zld **(a)**, Grh **(e)**, or Twi **(i)** when expressed at different levels.
**b,f,j,** Boxplots of ATAC-seq signal at class I-III regions for Zld **(b)**, Grh **(f)**, or Twi **(j)** when expressed at different levels.
**c,g,k,** Distribution of redefined class II and III peaks when redefined using ChIP-seq and ATAC-seq data for each concentration of CuSO~4~ for Zld **(c)**, Grh **(g)**, or Twi **(k)**.
Y-axis indicates the total number of peaks called at each level of induction.
Colors indicate the proportion of those peaks that are defined as class II, or III for each level of induction.
**d,h,l,** Venn diagrams showing overlap of peaks called at medium or high expression levels for Zld **(d)**, Grh **(h)**, or Twi **(l)**.
Expression was induced for 48 hours.

**Figure 6: The DNA-binding domain is not sufficient for pioneer-factor function. a,b,** Genome browser tracks showing example of class III regions where 48 hours of expression of the DBD alone is insufficient for robust binding and chromatin opening.
Adjacent class I regions are shown where the DBD alone can bind.
Tracks shown for Zld **(a)** and Grh **(b)**.
Class III regions are highlighted with gray shading.
**c,d,** Metaplots comparing ChIP-seq signal for full-length protein or DBD alone at class I, II and III regions for Zld **(c)** or Grh **(d)**.
CuSO~4~ levels used for induction are indicated for each DBD.
(1000 µM CuSO~4~ was used for full-length Zld. 100 µM CuSO~4~ was used for full-length Grh.) **e-j,** Volcano plots for ATAC-seq data upon expression of full-length Zld **(e)**, Zld DBD at low concentration **(f)**, Zld DBD at high concentration **(g),** full-length Grh **(h)**, Grh DBD at low concentration **(i)**, or Grh DBD at high concentration **(j)**.
Regions bound by full-length Zld are indicated in blue.
Regions bound by full-length -Grh are indicated in orange.
**k,** Model for the regulation of PFs.
A high level of chromatin occupancy is required for PFs to initiate chromatin opening.
Chromatin occupancy can be regulated by protein-intrinsic features such as motif affinity and protein domains, and by cell-intrinsic properties such as histone modifications, PF concentration, and cell-type specific cofactors (outlined in blue).

**Extended Data Figure 1: Stable cell lines allow inducible expression of transcription factors at physiological concentrations. a,** Schematic of generation of stable cell lines and induction of protein expression.
**b,** mRNA levels of *zld* and *grh* in S2 cells.
Top histogram shows the distribution of mRNA levels for all *Drosophila* genes.
Vertical dashed line indicates a log~2~ RPKM value of 0 as a threshold for considering a gene to be expressed.
**c-d,** Immunoblots showing titration of Zld **(c)** or Grh **(d)** protein levels in stable cell lines.
Two independently generated cell lines are shown and compared to 2-3 hours (H) embryos.
60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hours embryos.
Black arrowheads indicate Zld **(c)** or Grh **(d)**.
Gray arrowheads indicate background bands used to assess loading.
**e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells. ATAC-seq signal in wild-type cells is shown for reference.

@@

**Extended Data Figure 8: Analysis of chromatin and motif content at class I-VI regions. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
Tubulin levels are shown as a loading control.
**b,** Heatmap showing specificity of anti-H327me3 antibody in CUT&RUN reactions.
A panel of barcoded spike-in nucleosomes bearing different modifications was added to each CUT&RUN reaction (see methods).
For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.
**c-e,** Heatmaps showing the levels of different chromatin marks in class I-VI regions for Zld **(c)**, Grh **(d)** or Twi **(e)**.
**f-h,** Enrichment of known motifs in class I-VI sites for Zld **(f)**, Grh **(g)** or Twi **(h)**.
Left heatmap shows normalized motif rank within each class, with 1 being more enriched and 0 being less enriched.
Right heatmap shows the normalized expression (log~2~ RPKM) in RNA-seq datasets from each of the tissues that was analyzed.

**Extended Data Figure 9: Expression of Zld, Grh, or Twi at high protein levels results in chromatin opening at a small number of novel binding sites. a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld **(d)**, Grh **(e)**, or Twi **(f)** when expressed at varying concentrations.
**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.
**j,** Heatmap showing z-score normalized, anti-Zld CUT&RUN data generated from either wild-type neural stem cells (Type II neuroblasts) or neural stem cells over-expressing Zld.
Heatmaps are divided to show those peaks detected in wild-type vs. novel peaks that were only observed with over-expression of Zld.
ATAC-seq data from wild-type neural stem cells is also shown.
**i,** Heatmap of adjusted p-values for the top *de novo* motifs enriched in either the endogenous or novel (Zld overexpression) Zld binding sites.

@@

**Supplementary table 2: Previously published genomics datasets that were analyzed in this work.** This table provides the target of immunoprecipitation (for ChIP-seq datasets), the cell type and/or developmental stage, the Gene Expression Omnibus accession number (GSE), and the PubMed ID for the study in which the dataset was originally published.

### 2023-10-20 manuscript/supplemental_figures.qmd
update analysis and MS based on reviewer comments

**Stable cell lines allow inducible expression of transcription factors at physiological concentrations. a,** Schematic of generation of stable cell lines and induction of protein expression.
**b,** mRNA levels of *zld* and *grh* in S2 cells.
Top histogram shows the distribution of mRNA levels for all *Drosophila* genes.
Vertical dashed line indicates a log~2~ RPKM value of 0 as a threshold for considering a gene to be expressed.
**c-d,** Immunoblots showing titration of Zld **(c)** or Grh **(d)** protein levels in stable cell lines.
Two independently generated cell lines are shown and compared to 2-3 hours (H) embryos.
60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hours embryos.
Black arrowheads indicate Zld **(c)** or Grh **(d)**.
Gray arrowheads indicate background bands used to assess loading.
**e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells.
:::

@@

**Loss of H3K27me3 in tazemetostat-treated cells. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
Tubulin levels are shown as a loading control.
**b,** Heatmap showing specificity of anti-H327me3 antibody in CUT&RUN reactions.
A panel of barcoded spike-in nucleosomes bearing different modifications was added to each CUT&RUN reaction (see methods).
For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.
:::

@@

**Expression of Zld, Grh, or Twi at supraphysiological levels results in chromatin opening at a small number of novel binding sites. a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld **(d)**, Grh **(e)**, or Twi **(f)** when expressed at varying concentrations.
**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.
:::

>>>>

**Stable cell lines allow inducible expression of transcription factors at physiological concentrations. a,** Schematic of generation of stable cell lines and induction of protein expression.
**b,** mRNA levels of *zld* and *grh* in S2 cells.
Top histogram shows the distribution of mRNA levels for all *Drosophila* genes.
Vertical dashed line indicates a log~2~ RPKM value of 0 as a threshold for considering a gene to be expressed.
**c-d,** Immunoblots showing titration of Zld **(c)** or Grh **(d)** protein levels in stable cell lines.
Two independently generated cell lines are shown and compared to 2-3 hours (H) embryos.
60,000 cells were loaded in each well, which is equivalent to the approximately 60,000 nuclei present in 10 2-3 hours embryos.
Black arrowheads indicate Zld **(c)** or Grh **(d)**.
Gray arrowheads indicate background bands used to assess loading.
**e-g,** Heatmaps comparing Zld **(e)**, Grh **(f)**, or Twi **(g)** ChIP-seq signal to control experiments in which anti-Zld, anti-Grh, anti-Twi, or IgG antibodies were used to perform immunopreciptation in wild-type (WT) cells. ATAC-seq signal in wild-type cells is shown for reference.
:::

@@

**Analysis of chromatin and motif content at class I-VI regions. a,** Immunoblot showing H3K27me3 levels in two replicates of DMSO- or tazemetostat-treated cells.
Tubulin levels are shown as a loading control.
**b,** Heatmap showing specificity of anti-H327me3 antibody in CUT&RUN reactions.
A panel of barcoded spike-in nucleosomes bearing different modifications was added to each CUT&RUN reaction (see methods).
For each sample, the heatmap displays the percentage of barcode reads for each sample and histone modification relative to the total number of barcode reads for all modifications.
**c-e,** Heatmaps showing the levels of different chromatin marks in class I-VI regions for Zld **(c)**, Grh **(d)** or Twi **(e)**.
**f-h,** Enrichment of known motifs in class I-VI sites for Zld **(f)**, Grh **(g)** or Twi **(h)**. Left heatmap shows normalized motif rank within each class, with 1 being more enriched and 0 being less enriched. Right heatmap shows the normalized expression (log~2~ RPKM) in RNA-seq datasets from each of the tissues that was analyzed.
:::

@@

**Expression of Zld, Grh, or Twi at high protein levels results in chromatin opening at a small number of novel binding sites. a-c,** Immunoblots showing Zld **(a)**, Grh **(b)** or Twi **(c)** protein levels when stable cell lines are induced using different concentrations of CuSO~4~.
**d-f,** Bar plots showing the percentage of previously defined class I-VI binding sites that are bound by Zld **(d)**, Grh **(e)**, or Twi **(f)** when expressed at varying concentrations.
**g-i,** Bar plots showing the percentage of previously defined class I-VI binding sites that overlap an ATAC-seq peak when Zld **(g)**, Grh **(h)**, or Twi **(i)** are expressed at varying concentrations.
**j,** Heatmap showing z-score normalized, anti-Zld CUT&RUN data generated from either wild-type neural stem cells (Type II neuroblasts) or neural stem cells over-expressing Zld. Heatmaps are divided to show those peaks detected in wild-type vs. novel peaks that were only observed with over-expression of Zld. ATAC-seq data from wild-type neural stem cells is also shown.
**i,** Heatmap of adjusted p-values for the top *de novo* motifs enriched in either the endogenous or novel (Zld overexpression) Zld binding sites.
:::
